\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-19 nobuild/jss_revision/response.tex

\usepackage{geometry}
\geometry{margin=1.5in}

\begin{document}

{\bfseries
\begin{center}
  Manuscript JSS2717\\
 ``sparseHessianFD: An R Package for Estimating Sparse
 Hessian Matrices''\\
 \vspace{\baselineskip}
 Response to reviewers
\end{center}
}
\vspace{\baselineskip}

I would like to express my sincere appreciation for the time and effort that
the reviewer invested in the review.  It is obvious to me that the
reviewer went through the paper and code in great detail.  In
particular, I want to thank the reviewer for bringing the complex step
method to my attention.  I \emph{never} use complex numbers or
functions in my own work, so I typically ignore anything with the word
"complex" in it.  But this review prompted me to learn something new.
That is always fun.

I am happy to write that I was able to incorporate a large majority of
the reviewer's suggestions in the revised paper and package. For the
others, I hope that the editor and reviewer will accept my
explanations and excuses.  Of course, if there are any strong
objections to how I responded to the reviewer's concerns, I am happy
to make further changes in advance of publication.

\vspace{\baselineskip}

@@

{\normalfont (error messages related to algorithm.sty removed).}

I don't remember if there is an easy way to distribute ``special'' sty files,
or if you must force the user to install them. If the sty is not really
critical it might be better to omit it.

\end{revQuote}

\begin{response}
I suspect that the error is a result of the \pkg{algorithm} package
not being present in the reviewer's LaTeX installation.  I would like
to continue to use this package,
because it does allow for nicely formatted algorithms.  I believe that this is
a relatively common package, and one that should be easily installed
through CTAN.  However, if there is a strong
objection, I may be able to find another way to format the algorithms.
\end{response}

@@

\item \begin{revQuote}
I am not familiar with reference classes, but for most users I think it
should not be important that the package uses them. However, as a user, I
find it disconcerting that sparseHessianFD() is used as a (constructor)
function (eg. p12, and in the example in ?sparseHessianFD) but is not
documented in the help with 'usage' and 'arguments' as is usual for
functions. There seems to be little guidance in the usual R places about how
reference classes should be documented. Perhaps you could seek some guidance
from the community on this point.
\end{revQuote}

\begin{response}

  Note to me:  issue here is that there is no documentation for
  \func{sparseHessianFD()} as a function.  Good idea to create a
  documentation page for that.

@@

\item \begin{revQuote}
It is pointed out (p2) that requirements are burdensome and emphasized
that the package is not appropriate for all uses. Also, the conditions 1-5
are fairly stringent. The reader is left with the impression that there may
be few applications. I think the value of the package could be promoted
better. Perhaps a small list of different types or applications of
hierarchical problems could be given, or an indication of other classes of
problems where the conditions would apply.

  \end{revQuote}

\begin{response}

Note to me:  reword last paragraph on p2 to make the requirements
sound a lot less burdensome.  Spin these to show that these are
easy,but acknowledge that supplying a gradient might be difficult.

Then, follow reviewer's advice and explain situations in which the
requirements might not be met.

Note that there is really no disadvantage to defining the sparsity pattern as
less sparse than it is (other than suboptimal computation). So, if the
hierarchical structure is known, but there are potentially independent
components in the margins, there is nothing wrong with assuming dense margins.

@@

\end{response}

\item \begin{revQuote}
Condition 5 might be relaxed to a local condition in a neighborhood of
the evaluation point, rather than a global condition, but I do not know if
that would have any practical value.
  \end{revQuote}

\begin{response}

  In order to gain any kind of computational advantage, the sparsity
  pattern has to be the same for the entire domain of the function. I
  think that qualifies as a global condition in the reviewer's
  terminology.

  There is a distinction between a structural zero and a numerical
  zero.  A structural zero is an element of the Hessian that is zero
  for all possible variables.  Because that element is always zero, we
  do not have to reserve memory for it in the sparse matrix object.
  A Hessian element might be a numerical zero when evaluated at some
  values, but not others.

  Now, suppose we allowed the sparsity pattern to change when a zero
  element became non-zero.  At the moment, not only would we need to
  allocate memory for the new non-zero element, but we would need to
  rearrange all of the other elements to put the new element in the
  right place.  The efficiency of sparse matrix operations comes from
  how the non-zero elements are ordered and indexed.

Note to me:  somewhere in the paper, clarify the difference between
structural and numerical zeros, and explain why that matters.

@@

\end{response}

\item \begin{revQuote}
Section 2.1 p7 I think could be made more easily readable by adding a few
more hints about dimension, and more care about use of the term 
"coefficients vector", for example:

  - 'continuous covariates $x_i$,' -> 'continuous covariates $x_i \in
\Real{k}$,'

  - 'heterogeneous coefficient vector $\beta_i$' -> 'heterogeneous
coefficient vector $\beta_i \in \Real{k}$'

@@

\begin{verbatim}
print(obj$fn(P) - binary.f(P, 
      data=binary, priors=priors, order.row=order.row), digits=20)
[1] 0

print(max(abs(obj$gr(P) - binary.grad(P, 
         data=binary, priors=priors,
order.row=order.row))),digits=20)
[1] 0

\end{verbatim}

It is not clear to me whether obj\$fn() and obj\$gr() use code as in the true
functions or a modified version using sparse techniques. Some further
clarification would be helpful.

### 2016-12-20 nobuild/jss_revision/response.tex

\item \begin{revQuote}
It is pointed out (p2) that requirements are burdensome and emphasized
that the package is not appropriate for all uses. Also, the conditions 1-5
are fairly stringent. The reader is left with the impression that there may
be few applications. I think the value of the package could be promoted
better. Perhaps a small list of different types or applications of
hierarchical problems could be given, or an indication of other classes of
problems where the conditions would apply.

  \end{revQuote}

  \begin{response}

I included this section primarily as a way to avoid overselling the
method, but the point that the reader might be scared away by this list of
restrictions is well-taken.  In the revision, I removed the list of
conditions, and wove them into the text.  The result takes a more
positive tone that in the original submission.

@@

\end{response}

\item \begin{revQuote}
Condition 5 might be relaxed to a local condition in a neighborhood of
the evaluation point, rather than a global condition, but I do not know if
that would have any practical value.
  \end{revQuote}

\begin{response}

  The list of conditions is no longer in the paper, but to this point,
  I added a defintion of a ``structural zero'' in the third
  paragraph. The sparsity pattern represents the structure of the
  Hessian, not its value, so any element that \emph{could} be zero,
  but does not have to be zero, is not a structural zero.  Thus, the
  condition is global and not local.

  Note that other than
  a potential loss of efficiency, there is no disadvantage to erring
  on the side of an element being non-zero.  
\end{response}

### 2016-12-20 nobuild/jss_revision/sparseHessianFD.Rnw
revision

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=FALSE)
options(replace.assign=TRUE, width=77, prompt="R> ")
@

@@

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of non-zero elements in the
Hessian is small.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

@@

The package defines the \class{sparseHessianFD} class, whose initializer
requires the user to provide functions that compute the objective
function and the gradient, as well as the sparsity pattern of the
Hessian.  Some users may find those requirements burdensome, and we
emphasize that there may be some applications for
which \pkg{sparseHessianFD} is not an appropriate package to use. To
extract the maximum benefit from using \pkg{sparseHessianFD}, we need
to accept some conditions and assumptions.

\begin{enumerate}
  \item Preferred alternatives to computing the Hessian are not
  available.  Finite differencing is not generally a ``first choice''
  method.  Deriving a gradient or Hessian symbolically, and writing a
  subroutine to compute it, will give an exact answer, but might be
  tedious or difficult to implement. Algorithmic differentiation (AD) is
  probably the most efficient method, but requires specialized
  libraries that, at this moment, are not yet broadly available in \proglang{R}.
 \pkg{sparseHessianFD} makes the most sense when the gradient is easy to compute, but the Hessian is
  not.
  \item The application can tolerate the approximation error in the Hessian
  that comes with finite differencing methods.
\item The objective function $f(x)$ is twice differentiable, and can be
  computed ``quickly and easily,'' even for a large number of
  variables.  We leave the definition of ``quickly and easily''
  intentionally murky, since no method of differentiation can overcome
  pathologies in a function that itself is hard to compute.
\item The gradient can
  be computed quickly, easily and \emph{exactly} (within machine
  precision).   We do not recommend using finite differenced gradients
  when computing finite differenced Hessians.  The
  approximation errors will be compounded, and the time
  complexity of computing a gradient grows with the number of
  variables.
\item The sparsity pattern is known in advance, and does not depend
    on the values of the variables.
\end{enumerate}

Our perspective on the requirements to use \pkg{sparseHessianFD} is
that deriving a vector of first
derivatives, and writing \proglang{R} functions to compute them, is a
lot easier than doing the same for a matrix of second derivatives, and
more accurate than computing second-order approximations from the
objective function. Even when we have derived the Hessian
symbolically, in practice it may still be faster to estimate the Hessian
using \pkg{sparseHessianFD} than coding it directly.   These are the
situations in which \pkg{sparseHessianFD} adds the most value to the
statistician's toolbox.  If AD software is available to compute the gradient, then
it is probably available for sparse Hessians as well, and
\pkg{sparseHessianFD} would not be needed.  Without AD-computed or
symbolically derived Hessians, \pkg{sparseHessianFD} is the best
available option.

The only functions and methods the end user should need to use are the
\class{sparseHessianFD} initializer, methods that return the Hessian
in a sparse compressed format, and perhaps some utility functions that
simplify the construction of the sparsity pattern.  The class also
defines methods that partition
the variables into groups that can be perturbed together in a finite
differencing step, and recovers the elements of the Hessian via linear
substitution.  Those methods perform most of the work, but should be
invisible to the user.

>>>>

<<setup1, echo=FALSE, cache=FALSE>>=
suppressPackageStartupMessages(require(Matrix))
knitr::render_sweave()
knitr::opts_chunk$set(prompt=TRUE, cache=FALSE) %$
options(replace.assign=TRUE, width=77, prompt="R> ")
@

@@

However, for many problems, the Hessian is
\emph{sparse}, meaning that the proportion of  ``structural''  zeros
(Hessian matrix elements that are always zero, regardless of the value at which
function is estimated) is high.  Consider a log
posterior density in a Bayesian hierarchical
model.  If the outcomes across units are conditionally
independent, the cross-partial derivatives of heterogeneous variables
across units are zero.  As the number of units
increases, the size of the Hessian still grows quadratically, but the number
of \emph{non-zero} elements grows only linearly; the Hessian
becomes increasingly sparse.  The row and column indices of the
non-zero elements comprise the \emph{sparsity pattern} of the Hessian,
and are typically known in advance, before computing the values of
those elements.  \proglang{R} packages such as \pkg{trustOptim}
\citep{R_trustOptim}, \pkg{sparseMVN} \citep{R_sparseMVN} and
\pkg{ipoptr} \citep{R_ipoptr} have the capability to accept Hessians
in a compressed sparse format.

@@

The package defines the \class{sparseHessianFD} class, whose initializer
requires the user to provide functions that compute an objective
function, its gradient, and the sparsity pattern of its Hessian
matrix. The gradient should be computed exactly (within machine
precision), as opposed to being numerically estimated (e.g., by finite
differencing). The sparsity pattern (e.g., location of structural
zeros) must be known in advance, and cannot vary across the domain of the
objective function. The only functions and methods of the class that
the end user should need to use are the
initializer, methods that return the Hessian
in a sparse compressed format, and perhaps some utility functions that
simplify the construction of the sparsity pattern.  The class also
defines methods that partition the variables into groups that can be perturbed together in a finite
differencing step, and recovers the elements of the Hessian via linear
substitution.  Those methods perform most of the work, but should be
invisible to the user.

As with any computing method or algorithm, there are boundaries around
the space of applications for which \pkg{sparseHessianFD} is the right
tool for the job. In general, finite differencing is not a ``first
choice'' method because the result is not exact, so
\pkg{sparseHessianFD} should not be used when the application cannot
tolerate any error, no matter how small.  Also, we admit that some
users might balk at having to provide an exact gradient, even though
the Hessian will be estimated numerically. However, deriving a vector of first
derivatives, and writing \proglang{R} functions to compute them, is a
lot easier than doing the same for a matrix of second derivatives, and
more accurate than computing second-order approximations from the
objective function. Even when we have derived the Hessian
symbolically, in practice it may still be faster to estimate the Hessian
using \pkg{sparseHessianFD} than coding it directly.   These are the
situations in which \pkg{sparseHessianFD} adds the most value to the
statistician's toolbox.
